---
title: "Mapping in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn to use RStudio to pull data from TidyCensus, 
  compare map design, and visualize accompanying data.
---

```{r setup, include=FALSE}
library(learnr)
library(httr)
library(jsonlite)
library(tidyverse)
library(sf)
library(tidycensus)
library(rempsyc)
library(biscale)
library(cowplot)
```


## Welcome

In this tutorial, you will learn how the basic skills of mapping spatial 
relationships in Rstudio including:

* The basics of RStudio for this application, including tidy data and RMarkdowns. 
* How to pull data from an Application Programming Interface (API)
* How to combine datasets based on their spatial relationships
* How to conduct exploratory analyses
* How to visualize spatial data 

### Setup

To practice these skills, we will answer the question, Are there any sociodemographic 
patterns in public Electric Vehicle (EV) infrastructure  for Davidson County, TN? 
We will access EV charging station data from the [National Renewable Energy Laboratory (NREL)](https://developer.nrel.gov/docs/transportation/alt-fuel-stations-v1/all/), 
and sociodemographic data from the [American Community Survey](https://www.census.gov/programs-surveys/acs/data.html). 

I've preloaded the packages for this tutorial with the following: 

```{r load packages, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse) # loads dplyr, ggplot2, and others
library(tidycensus) # provides access to ACS data
library(httr) # for API requests
library(jsonlite) # for API requests
library(sf) # for mapping shapefiles
library(cowplot) # for combining plots
library(rempsyc) # for statistical tests
library(biscale) # for bivariate mapping
```


## Download data from an API

### EV Charging Data

**_NOTE:_**  Before we get started, you'll have to [register for an API key](https://developer.nrel.gov/signup/), which will allow you to pull your
very own data requests. API keys should not be shared. 

Once you've got your key, go ahead and paste it in the next chunk.
```{r load API key, echo=TRUE}
my_api_key <- "" # paste key between quotations
```

```{r load my API, include=FALSE}

my_api_key <- "hRugwQrdFhJm7IYUiOg37ghpYgjXBoOecbh4gOxn"
```

APIs are fabulous alternatives to manually downloading data, allowing researchers
to ensure their data analysis is replicable, from download to publication. In our
case, we will be submitting a "GET" request, which signals the API that you'd
like some of its data. Our GET request will be made up of a few parts:

* Supplying a URL for the API
* Limiting our request to certain query parameters
* Checking that our request was successful
* Converting our request from JSON to text

First, let's supply our URL and query parameters. When searching for
publicly-available datasets, APIs are often made available with accompanying
documentation. For our EV data, NREL kindly supplies its [documentation](https://developer.nrel.gov/docs/transportation/alt-fuel-stations-v1/all/)
on its site.

```{r creating API request, echo=TRUE, message=FALSE, warning=FALSE}
# set the primary URL for the API
nrel_url <- "https://developer.nrel.gov/api/alt-fuel-stations/v1.json?"

# set the target state for the data retrieval, in this case, TN
state <- "TN"

# paste together the full string to retrieve the data
nrel_api_url <- paste0(nrel_url, "api_key=", my_api_key, "&state=", state, "&limit=all")

# use the full api url to get the data
EV_data <- GET(nrel_api_url)
```

Now, let's make sure that request was successful. 
```{r GET request status, echo=TRUE, warning=FALSE}
EV_data
```

This summary tells us the URL we sent the request to, the timestamp, the type of 
content, and the status. Status is important, because it let's us know if there
were any problems in our query. 200 means your request was successful, but in the
event of a different code, here's a [list](https://www.restapitutorial.com/httpstatuscodes.html) 
of other codes you may encounter. 

Let's convert our JSON request to something more legible in R. In this chunk, we're 
using the function `fromJSON()` to convert the raw data to character data. This dataframe 
is actually several lists in a dataframe, let's extract only the data by subsetting 
the dataframe we're interested in, entitled `fuel_stations`.

```{r converting from JSON, echo=TRUE, message=FALSE, warning=FALSE}
EV_data <- jsonlite::fromJSON(rawToChar(EV_data$content))

EV_data <- EV_data[["fuel_stations"]]
```

Great, this is more aligned with something we might be more accustomed to. Our 
public EV charging station data for the state of TN has `r format(nrow(EV_data), big.mark = ",")` 
entries, which we should probably trim down to only represent Davidson County, TN. 

First, we give our new dataframe a name, I call it `nash_EV`, so that I
remember it's a subset of our TN dataset. We then call our dataframe of interest,
`EV_data`, and we pipe it into our next command (pipes are used to connect and 
simplify your commands; they basically tell R to do this "and then" that). 
Next, we'll filter our data for `city` (Nashville, TN), `fuel_type` (Electric),
and `access_code` (public). 

We'll also go ahead and simplify our dataframe by selecting for the variables we're 
most interested in, `status_code`, `access_code`, `fuel_type_code`, `owner_type_code`,`latitude`,
and `longitude`. 

```{r filtering for Nashville, echo=TRUE, message=FALSE, warning=FALSE}
nash_EV <- EV_data %>%
            filter(city == "Nashville" & 
                   fuel_type_code == "ELEC" &
                   access_code == "public") %>% 
            dplyr::select(status_code, access_code, fuel_type_code, owner_type_code,
                   latitude, longitude)

#take a look
glimpse(nash_EV)  
```

Shall we map it? Because we've got latitude and longitude, it's relatively easy
to represent this data spatially, as points. 

**_NOTE:_**  Datums, projections and coordinate systems matter! Datums model the 
Earth, so that they can be used for coordinate systems. coordinate systems provide 
numerical information that allows you to position your data on Earth. Meanwhile, 
Projections are how you, the mapmaker, translate the curved surface of the earth 
to a flat surface.

For this project, I'll use the NAD83 datum, because it's what most federal agencies 
use. While I could opt to use a State Plane Coordinate System, I'm not doing any
overly complex distance calculations, so WGS84 should be just fine. I encourage 
you to read more about projections and coordinate systems 
[here](https://mgimond.github.io/Spatial/chp09_0.html)


We'll use the package, `sf` to convert our `nash_EV` dataframe into an sf object. 
The `st_as_sf()` function is our powerhouse here, and we supply it with the `nash_EV`
dataframe, the coordinate columns, the coordinate reference system (in this case
it is NAD83), please see this [resource](https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf)
for more information. 

```{r making EV data spatial, echo=TRUE, fig.height=4, fig.width=7.5, message=FALSE, warning=FALSE}

# create shapefile
ev_sf <- nash_EV %>%
  st_as_sf(
    coords = c(x = "longitude", y = "latitude"),
    crs = "EPSG:4269" #define CRS for NAD83
    )

# check that crs worked
st_crs(ev_sf)

# plot!
ggplot(ev_sf) +
    geom_sf() 
```

Phew, we did it! Unfortunately though, we need a bit more information to call 
this an effective map. Let's add in some sociodemographics, next.

### ACS Data

A [2022 paper](https://doi.org/10.1016/j.tej.2022.107096 ) published by Khan et. al.,
assessed EV charging stations in New York City, suggesting that EV charger locations 
may be correlated to sociodemographic variables. Interestingly, median household 
income, percentage of white individuals, and presence of a highway were positively 
correlated with the presence of an EV charging station. Meanwhile, the percentage 
of Black individuals in a zip code were negatively correlated.

Let's see if these patterns hold true for the city of Nashville. 

We'll need race/ethnicity variables and household income information from the 
census. Let's use the `TidyCensus` package to download our data directly from 
the Bureau's API.

**_NOTE:_**  You'll have to [register for a Census Bureau API key](https://api.census.gov/data/key_signup.html) first, which will allow you 
to pull your very own data requests. API keys should not be shared. You can use the 
`census_api_key()` function to store your key for the session. 

For this session, we'll use ACS data from the US Census Bureau, using 5-year estimates. 
We can load all of the variables and their unique codes using the function 
`load_variables()`. We'll specify `acs5` as the dataset, and cache it, so things 
move a bit faster. 

Next, we'll create a query for our data using the function, `get_acs()`. We'll provide it 
the following parameters: 

* `geography` specifies the level of geography you'd like. We'll use tract for this
session, but there are several [other levels](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html) you can query. 
* `survey` specifies the type of ACS data you'd like. We'll use ACS 5-year estimates
* `variables` can be defined using the `acs_vars` dataframe we created. We'll create
a list for total population, median income, as well as race/ethnicity values.
* `state` will specify the US state of interest, in this case TN. 
* `county` will specify a county of interest, in this case Davidson. 
* `year` will specify the end date for the ACS estimate, in this case we'll use 
the 2016 to 2020 dataset.
* `geometry` will specify whether we want a shapefile or not, which in this case we do.
* `output` will specify the shape of the data, which in this case we'd like wide, 
rather than long. 

```{r TidyCensus Request, echo=TRUE, results = FALSE, message=FALSE, warning=FALSE}

# Run this line first: census_api_key("YOUR API KEY GOES HERE") 

acs_vars <- load_variables(2020, "acs5", cache = TRUE) # Get ACS variables

davidson_acs <- get_acs(geography = "tract", 
                survey = "acs5",        
                variables = c(total_pop = "B03002_001",
                              medincome = "B19013_001",
                              white_non_hisp = "B03002_003", 
                              black_non_hisp = "B03002_004", 
                              asian_non_hisp = "B03002_006", 
                              hisp = "B03002_012"), 
                state = "TN",
                county = "Davidson",
                year = 2020, 
                geometry = TRUE,
                output = "wide")
```

This created a multipolygon dataset, with `r format(nrow(davidson_acs))` rows, 
representing each census tract in the county and the Estimate and Margin of 
Error for each variable of interest. 
```{r}
glimpse(davidson_acs)
```


**_NOTE:_**  Because the ACS is based on estimates rather than official counts 
(like the decennial census), understanding the Margin of Error associated with 
each estimate is incredibly important; I'd suggest reading more [here](https://walker-data.com/tidycensus/articles/margins-of-error.html!)

Let's go ahead and standardize these counts, so we can compare percentages. We'll
use the `mutate()` function to create percentage variables for race/ethnicity, and 
we'll also select only the variables we need. 

```{r Tidycensus Formatting, echo=TRUE, message=FALSE, warning=FALSE}
davidson_tract <- davidson_acs %>% 
                  mutate(perc_nh_white = (white_non_hispE/total_popE)*100, 
                         perc_nh_black = (black_non_hispE/total_popE)*100,
                         perc_nh_asian = (asian_non_hispE/total_popE)*100, 
                         perc_hisp = (hispE/total_popE)*100,
                         medhhinc = medincomeE) %>%
      dplyr::select(GEOID, total_popE, perc_nh_white, perc_nh_asian, 
                    perc_nh_black, perc_hisp, medhhinc)

glimpse(davidson_tract)
```

Quick check to ensure the CRS is the same as our nashville EV charging infrastructure
```{r ACS CRS check, echo=TRUE, message=FALSE, warning=FALSE}
st_crs(davidson_tract)
```

Finally, let's conduct a few quick checks, just to ensure the data looks alright. 

First, let's create a quick plot of median household income. We'll use the `ggplot2` 
package for this tutorial, along with the `sf` package, but know there are loads of 
other mapping packages out there. 

To map our tract shapefile, we'll define the spatial dataframe being used in 
the `ggplot()` function, then add a `geom_sf()` call, with `medhhinc` in the 
`aes()` function. We can also clean up some of the labels. 

```{r Map MHI, echo=TRUE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7.5}
ggplot(data = davidson_tract) +
  geom_sf(aes(fill = medhhinc)) + 
  labs(title = "Median Household Income in Nashville, TN",
       fill = "Median Household \n Income (USD)",
       caption = "Source: 2016-2020 ACS, US Census Bureau") 
```

Those greyed tracts are interesting, we can take a look at the NAs in the data 
and compare them with this map to ensure that they're non-residential. We can use
a [tract map](http://maps.nashville.gov/webimages/MapGallery/PDFMaps/2010%20Census%20Tracts.pdf) 
of Nashville to double check. It looks like these tracts are the BNA airport, and 
a smaller airport/industrial area. 
```{r check for NA census tracts, echo=TRUE, message=FALSE, warning=FALSE}
davidson_tract %>% 
  filter(is.na(medhhinc)) %>% 
  dplyr::select(GEOID)
```


## Combining Spatial Data

To begin some exploratory analyses on the relationship between EV charging 
stations and sociodemographics, we need to summarize the EV locations to the 
tract level. 

We'll use the function `st_join()`, defining the points (`ev_sf`) and the 
polygons (`davidson_tract`), using the join type, `st_within()`. 

Following, we'll summarize the tract_count by using the `count()` function, 
specifying that we'd like a count for each unique `GEOID`. We'll also rename the 
default variable name (`n`) to `ev_count`. 

```{r Summarize EV locations, echo=TRUE, message=FALSE, warning=FALSE}
# Summarize EV locations
ev_in_tract <- st_join(ev_sf, davidson_tract, join = st_within)

# count EV stations per census tract
ev_tract_count <- count(as_tibble(ev_in_tract), GEOID) %>%
                  rename(ev_count = n) 

glimpse(ev_tract_count)
```

And we'll join! Use a `join()` function define the left dataframe, `davidson_tract` 
that we'll be joining to the `ev_tract_count`. Then, we use `mutate()` to fill the
replace the NAs in our `ev_count` column with zeroes. We also add a new column, 
`presence_stations`, which is a binary factor. Lastly, we remove any rows that 
have an NA for `medhhinc` (i.e., the airport/commercial spaces we saw above). 

```{r Join with tract, echo=TRUE, message=FALSE, warning=FALSE}
# join ev count with tract df, calc EV station density
tract_ev_sf <- left_join(davidson_tract, ev_tract_count) %>%
               mutate(ev_count = ifelse(is.na(ev_count), 0, ev_count),
                      presence_stations = factor(ifelse(ev_count > 0, 1, 0))) %>% 
               na.omit(medhhinc)

glimpse(tract_ev_sf)
```

Let's also do a quick map! Using the same approach as above, we'll layer two 
shapefiles (one point, one polygon). 
```{r Mapping the two, echo=TRUE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7.5}
ggplot() +
  geom_sf(data = davidson_tract) +
  geom_sf(data = ev_sf, fill = "black") +
  labs(title = "EV Charge Stations Located in Davidson County Tracts",
       caption = "Source: NREL + US Census") +  
  theme_bw()
```

## Exploratory Analysis
### Boxplots

Now that we've got the data summarized to the tract level, let's start by creating some 
boxplots.

To make several and compare at once, I'll save them as individual objects, and use 
the package `cowplot` to bind them all together. 

**_NOTE:_** Functions are the ideal approach to simplifying repetitive processes, 
but functions can be a bit complex. Read more [here](https://r4ds.hadley.nz/functions) if you're interested!

To make boxplots, we define the dataframe (`tract_ev_sf`) and pipe it into a `ggplot()`
object. We'll assign the aesthetics for the factor variable, `presence_stations`, 
and the continuous variable, `medhhinc`. Then, add a `geom_boxplot()` object, and 
specify a color of your choice. 

Repeat for other variables of interest, and plot the grid.

```{r boxplot facet, echo=TRUE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7.5}
a <- tract_ev_sf %>% 
      ggplot(aes(presence_stations, medhhinc)) + 
      geom_boxplot(fill = "#75A6A2")

b <- tract_ev_sf %>% 
      ggplot(aes(presence_stations, total_popE)) + 
      geom_boxplot(fill = "#75A6A2")

c <- tract_ev_sf %>% 
     ggplot(aes(presence_stations, perc_nh_white)) + 
    geom_boxplot(fill = "#75A6A2")

d <- tract_ev_sf %>% 
     ggplot(aes(presence_stations, perc_nh_asian)) + 
    geom_boxplot(fill = "#75A6A2")

e <- tract_ev_sf %>% 
      ggplot(aes(presence_stations, perc_nh_black)) + 
      geom_boxplot(fill = "#75A6A2")

f <- tract_ev_sf %>% 
      ggplot(aes(presence_stations, perc_hisp)) + 
      geom_boxplot(fill = "#75A6A2")

cowplot::plot_grid(a, b, c, d, e, f, ncol = 3)
```
Looks like there could be some interesting patterns between race/ethnicity and 
EV station locations. 

### Correlative patterns

Khan et al., used unpaired t-tests to assess the relationships between the 
presence of EV charging stations and sociodemographics, so let's start there. 
First, we'll use the `rempsyc` package for it's `nice_t_test()` function. Start 
by calling our dataframe, `tract_ev_sf`, then select the variables of interest, 
in this case `presence_stations`, and all of the columns from `total_popE` to 
`medhhinc`. Then, we use the `st_drop_geometry()`function to temporarily drop 
the spatial component of our dataframe. 

Lastly, we use the `nice_t_test()` function to define the columns of interest, 
and the `group`, which is `presence_stations`.

We can use the `nice_table()` function to create a prettier table. 

```{r Correlation Table, echo=TRUE, message=TRUE, warning=FALSE}
t.test.results <- tract_ev_sf %>% 
                  dplyr::select(presence_stations, total_popE:medhhinc) %>% 
                  st_drop_geometry() %>% 
                  nice_t_test(response = names(tract_ev_sf)[2:7],
                  group = "presence_stations")

nice_table(t.test.results)
```

It looks like there is some positive correlation between the presence of EV 
charging stations in more populated census tracts, as well as tracts with more 
Black individuals. Interestingly, there's a negative correlation with tracts 
containing more non-Hispanic white individuals. Let's see if we can map the two 
together.

## Maps
Finally, more maps! Although there's plenty of ways to map this information,
we'll focus on bivariate mapping, which is similar to a chloropleth, 
but allows us to visualize two variables (i.e., EV charging station counts and 
percentage of non-Hispanic white individuals) in the same map. 

To do so, we'll use the `biscale` package, and create a new dataframe `bi_white`, 
so that bivariate classes are assigned to the data using the `bi_class` function. 

We use `perc_nn_white` as our `x`, `ev_count` as our `y`, and fisher breaks for 
our `style`, and 3 colors for the `dim` option. 

**_NOTE:_**  Legend breaks vary by the story we're trying to tell. In this case, 
I chose Fisher's because it optimizes the classifications for a specific number 
of classes, which in this case is 3. There's a handy Book chapter that details 
the differences in legend classifications [here](https://geographicdata.science/book/notebooks/05_choropleth.html#). 

```{r Bivariate df, echo=TRUE, message=FALSE, warning=FALSE}
bi_white <- bi_class(tract_ev_sf, x = perc_nh_white, y = ev_count, 
                     style = "fisher", dim = 3)
```

The `biscale` package uses the `ggplot2` approach, which we're already familiar
with! We create a `ggplot()` object, add a `geom_sf` object, specifying our new
`bi_white` dataframe, our fill column, `bi_class`, and some other aesthetic 
details, such as line color and width.  

I chose the "Bluegill" palette for this approach, but there are several [others](https://cran.r-project.org/web/packages/biscale/vignettes/bivariate_palettes.html)
available. 

```{r Bivariate ggplot, echo=TRUE, message=FALSE, warning=FALSE}

white_map <- ggplot() +
             geom_sf(data = bi_white, aes(fill = bi_class), 
                     color = "white", size = 0.1, show.legend = FALSE) +
            bi_scale_fill(pal = "Bluegill", dim = 3) +
            bi_theme()
```

I also want to create a nice legend to accompany this, so I'll use the `bi_legend`
function to specify my palette, the number of colors, and the labels I'd like. 

```{r Bivariate legend, echo=TRUE, message=FALSE, warning=FALSE}
white_legend <- bi_legend(pal = "Bluegill",
                          dim = 3,
                          xlab = "  % White",
                          ylab = "Charging Stations",
                          size = 8)
```

The last step is adding the map and legend together, using the `ggdraw()` function. 

```{r add elements, echo=TRUE, message=FALSE, warning=FALSE, fig.height=4, fig.width=7.5}
whitePlot <- ggdraw() +
  draw_plot(white_map, 0, 0, 1, 1) +
  draw_plot(white_legend, 0.05, .6, 0.2, 0.2)

whitePlot
```

If we wanted to repeat this process, we could make bivariate maps for percentage 
of Black individuals and total population. Then we can combine all of the
figures into one final plot. 

```{r repeat for other vars, echo=TRUE, results = FALSE, message=FALSE, warning=FALSE}
## Black
bi_black <- bi_class(tract_ev_sf, x = perc_nh_black, y = ev_count, style = "fisher", dim = 3)

black_map <- ggplot() +
             geom_sf(data = bi_black, aes(fill = bi_class), 
                     color = "white", size = 0.1, show.legend = FALSE) +
            bi_scale_fill(pal = "Bluegill", dim = 3) +
            bi_theme()
  
black_legend <- bi_legend(pal = "Bluegill",
                          dim = 3,
                          xlab = "  % Black",
                          ylab = "Charging Stations",
                          size = 8)

blackPlot <- ggdraw() +
  draw_plot(black_map, 0, 0, 1, 1) +
  draw_plot(black_legend, 0.05, .6, 0.2, 0.2)

## Total Population
bi_pop <- bi_class(tract_ev_sf, x = total_popE, y = ev_count, style = "fisher", dim = 3)

pop_map <- ggplot() +
             geom_sf(data = bi_pop, aes(fill = bi_class), 
                     color = "white", size = 0.1, show.legend = FALSE) +
            bi_scale_fill(pal = "Bluegill", dim = 3) +
            bi_theme()
  
pop_legend <- bi_legend(pal = "Bluegill",
                          dim = 3,
                          xlab = "Total Pop",
                          ylab = "Charging Stations",
                          size = 8)

popPlot <- ggdraw() +
  draw_plot(pop_map, 0, 0, 1, 1) +
  draw_plot(pop_legend, 0.05, .6, 0.2, 0.2)
```

Plot together using cowplot
```{r Plot Together, echo=TRUE, results = FALSE, message=FALSE, warning=FALSE}
final_map <- cowplot::plot_grid(whitePlot, blackPlot, popPlot, nrow = 1, 
                   labels = c("% White Individuals", "% Black Individuals", 
                              "Total Population"))
```

Congratulations!! You've completed this Mapping in R Tutorial, and learned more 
about Nashville's transportation infrastructure along the way! 🌟🤩

## Additional Resources
If you're interested learning more, here are some extra resources that are all freely
available to you online:

* [R for Data Science 2](https://r4ds.hadley.nz/)
* [TidyCensus Tutorials](https://walker-data.com/tidycensus/index.html)
* [Hands on Programming with R](https://rstudio-education.github.io/hopr/)
* [Geocomputation with R](https://r.geocompx.org/)
* [Simple Features for R](https://r-spatial.github.io/sf/index.html)
* Fun Blog: [Julia Silge's Website](https://juliasilge.com/blog/)

### Contact me!
As always, please feel free to email me with any questions at 
[mariah.d.caballero@vanderbilt.edu](mailto:mariah.d.caballero@vanderbilt.edu)
